Modelling M/EEG data with Bayesian nonparametric multilevel models

### 2025-02-06 brms_meeg.qmd
now modelling actual decoding results (instead of simulated ones)

---
title: "Modelling M/EEG data with Bayesian nonparametric multilevel models"
shorttitle: "Bayesian M/EEG modelling"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysoiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive process and how they unfold over time. Typically, we are interested in testing whether such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset and offset of effects (e.g., differences across conditions). Here, we introduce a novel *model-based approch* for analysing one-dimensional M/EEG timeseries such as ERPs or decoding timecourses and their differences across conditions. This approach relies on Bayesian nonparametric multilevel modelling (multilevel generalised additive models or Gaussian processes), which outputs posterior probabililities of the effect being above chance at every timestep/voxel, while naturally taking into account the temporal dependencies and between-subject variability present in such data."
keywords: [EEG, MEG, generalised additive models, gaussian processes, mixed-effects models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography: meeg_modelling.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(permuco)
library(scales)
library(scico)
library(knitr)
library(brms)

@@

# setting a default ggplot2 theme
# theme_set(theme_light(base_size = 10, base_family = "CMU Sans Serif") )
theme_set(theme_light(base_size = 10, base_family = "Open Sans") )
```

@@

$$
\begin{aligned} 
y_{i} &\sim \mathrm{EF}\left(\mu_{i}, \phi\right)\\
g\left(\mu_i\right) &= \eta_{i} = A_{i} + \mathbf{X}_{i} \gamma + \sum_{j=1}^{J} f_{j}\left(x_{ij}\right)
\end{aligned}
$$

where $g(\cdot)$ is the link function, $A$ is an offset, $\mathbf{X}_{i}$ is the $i$ th row of a parametric model matrix, $\gamma$ is a vector of parameters for the parametric terms, $f_{j}$ is a smooth function of covariate $x_{j}$, $y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)$ denotes that the observations $y_{i}$ are distributed as some member of the exponential family of distributions (e.g., Gaussian, Gamma, Beta, Poisson) with mean $\mu_{i}$ and scale parameter $\phi$.

The smooth functions $f_{j}$ are represented in the model via penalised splines basis expansions of the covariates, that are a weighted sum of basis functions:

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$ th basis function $b_{j k}()$ evaluated at the covariate value $x_{ij}$ for the $j$ th smooth function $f_{j}$...

@@

A Gaussian process (GP) is a stochastic process that defines the distribution over a collection of random variables indexed by a continuous variable, that is $\{f(t): t \in \mathcal{T}\}$ for some index set $\mathcal{T}$ [@riutort-mayol_practical_2023; @rasmussen2005].

Note that nonparametric does not mean there aren't parameters, it means that there are infinitely many parameters.

From [brms documentation](https://www.rdocumentation.org/packages/brms/versions/2.22.0/topics/gp): A GP is a stochastic process, which describes the relation between one or more predictors $x=\left(x_1, \ldots, x_d\right)$ and a response $f(x)$, where $d$ is the number of predictors. A GP is the generalization of the multivariate normal distribution to an infinite number of dimensions. Thus, it can be interpreted as a prior over functions. The values of $f()$ at any finite set of locations are jointly multivariate normal, with a covariance matrix defined by the covariance kernel $k_p\left(x_i, x_j\right)$, where $p$ is the vector of parameters of the GP:

$$
\left(f\left(x_1\right), \ldots f\left(x_n\right) \sim \operatorname{MVN}\left(0,\left(k_p\left(x_i, x_j\right)\right)_{i, j=1}^n\right)\right.
$$

The smoothness and general behaviour of the function $f$ depends only on the choice of covariance kernel, see <https://mc-stan.org/docs/functions-reference/matrix_operations.html#gaussian-process-covariance-functions>. It ensures that values that are close together in the input space will be mapped to similar output values... From this perspective, $f$ is a realisation of an infinite dimensional normal distribution:

@@

where $C$ is a covariance kernel with hyperparameters $\lambda$ that defines the covariance between two function values $f\left(t_1\right)$ and $f\left(t_2\right)$ for two time points $t_1$ and $t_2$ [@rasmussen2005]. Similar to the different choices of the basis function for splines, different choices of the covariance kernel lead to different GPs. In this article, we consider the squared-exponential (a.k.a. Gaussian or radial basis function) kernel, which computes the squared distance between points and converts it into a measure of similarity. It is defined as:

$$
C(\lambda):=C\left(t_1, t_2, \sigma, \gamma\right):=\sigma^2 \exp \left(-\frac{\left(t_1-t_2\right)^2}{2 \gamma^2}\right)
$$

with hyperparameters $\lambda=(\sigma, \gamma)$, expressing the overall scale of GP and the length-scale, respectively [@rasmussen2005]. The advantages of this kernel are that it is computationally efficient and (infinitely) smooth making it a reasonable choice for the purposes of the present article. Here again, $\lambda$ hyperparameters are estimated from the data, along with all other model parameters.

@@

Following the approach used by @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see details and code in @yeung2004]...

@@

Copy-pasted from @rousselet_using_2025: "The signal was a truncated Gaussian defining an objective onset at 160 ms, a maximum at 250 ms, and an offset at 342 ms. It was created by getting the probability density function of a standard normal with 93 points from x= 1.5 to 1.5, rescaled in amplitude between 0 and 1 and padded with 79 zeros to the left and to the right, to form a time series of 251 points, from 0 to 500 ms, in steps of 2 ms (500 Hz sampling rate)."

@@

@@

```{r, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Slope for the difference between conditions according to the GAM (left) or the GP (right)."}
# defining a function to plot posterior slope
plot_post_slope <- function (model, data, eps = 0.1) {

@@

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = x, y = prob_ratio) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_line() +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 15,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
    labs(
        x = "Time (ms)",
        y = expression(Pr(beta>0~"|"~data, model) / (1 - Pr(beta>0~"|"~data, model) ) )
        )

@@

# plotting it
gp_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = x, y = prob_ratio) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_line() +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 15,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
    labs(
        x = "Time (ms)",
        y = expression(Pr(beta>0~"|"~data, model) / (1 - Pr(beta>0~"|"~data, model) ) )
        )

@@

@@

    # plotting it
    prob_y_above_0 %>%
        mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
        ggplot(aes(x = time, y = prob_ratio) ) +
        geom_hline(yintercept = 1, linetype = "dashed") +
        geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
        geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
        geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
        geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
        geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
        geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
        geom_vline(xintercept = 0, linetype = 2) +
        geom_line() +
        geom_point(
            data = . %>% dplyr::filter(!is.na(above_thres) ),
            aes(y = threshold),
            colour = "darkgreen",
            shape = 15,
            size = 3,
            na.rm = TRUE
            ) +
        scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
        labs(
            x = "Time (s)",
            y = expression(Pr(AUC>chance~"|"~data, model) / (1 - Pr(AUC>chance~"|"~data, model) ) )
            )

@@

```{r fig-sim-timegen, echo = FALSE, dpi = 300, fig.cap = "Exemplary (simulated) group-level average cross-temporal generalisation matrix of decoding performance (ROC AUC)."}
# number of timepoints
n_participants <- 20
# n_timepoints <- 240 # (1.2s at 200Hz)
n_timepoints <- 60 # (1.2s at 50Hz)
timepoints <- seq(from = -0.2, to = 1, length.out = n_timepoints)

@@

# plotting cross-temporal generalization matrix
timegen_data %>%
    summarise(auc = mean(auc), .by = c(train_time, test_time) ) %>%
    # pull(auc) %>% hist()
    ggplot(aes(x = train_time, y = test_time, fill = auc) ) +
    geom_tile() +
    geom_vline(xintercept = 0, linetype = 2, linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = 2, linewidth = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, linewidth = 0.5) +
    scale_x_continuous(expand = c(0, 0) ) +
    scale_y_continuous(expand = c(0, 0) ) + 
    coord_fixed() +
    scale_fill_scico(palette = "vik", midpoint = 0.5, name = "AUC") +
    # theme_light() +
    labs(x = "Testing time (s)", y = "Training time (s)")
```

Now, we want to test whether and when decoding performance is above chance level (0.5 for a binary decoding task). This one is computationally very intense...

```{r gam-timegen, message = FALSE}
# fitting a GAM with a 2D smooth function
timegen_gam <- brm(
    # 2D thin-plate spline (tp) to model smooth interactions between training and testing time
    auc ~ s(train_time, test_time, bs = "tp", k = 10),
    data = timegen_data,
    family = Beta(),
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/timegen_gam.rds"
    )
```

@@

```{r gp-timegen, eval = FALSE, message = FALSE}
# fitting a GP with two temporal dimensions
timegen_gp <- brm(
    auc ~ gp(train_time, test_time, k = 20),
    data = timegen_data,
    family = Beta(),
    control = list(adapt_delta = 0.95, max_treedepth = 20),
    iter = 2000,
    chains = 4,
    cores = 4,
    file = "models/timegen_gp.rds"
    )
```

@@

# checking timecourse of timegen_data's diagonal
timegen_data %>%
    filter(train_time == test_time) %>%
    summarise(auc = mean(auc), .by = train_time) %>%
    ggplot(aes(x = train_time, y = auc) ) +
    geom_hline(yintercept = 0.5, linetype = 2) +
    geom_line() +
    # theme_light() +
    labs(x = "Time (s)", y = "AUC")
```

@@

## Comparing the identified onsets to other approaches

Comparing performance with uncorrected massive test, classical correction procedures (e.g., Bonferroni, FDR, FWER), cluster-based permutation test, cluster-depth, change point, [TFCE](https://jaromilfrossard.github.io/permuco4brain/articles/tfce.html)... We used the `permuco` package [@frossard2021]...

```{r permutations-1d, echo = FALSE, fig.cap = "T-values timecourse with onset identified using the permuco package (clustermass method)."}
# centering AUC relative to chance level (0.5)
decoding_data_perm <- decoding_data %>%
    mutate(auc_centered = auc - 0.5) %>%
    dplyr::select(-auc)

@@

TO-DO: Prepare a wrapper `R` package and show how to call it in `Python` and integrate it with `MNE-Python` [@gramfort2013] pipelines...

@@

@@

$$
\text{TFCE} = \int_{h} e(h)^{E} h^{H} dh
$$

>>>>

---
title: "Modelling M/EEG data with Bayesian nonparametric multilevel models"
shorttitle: "Bayesian M/EEG modelling"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysoiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive process and how they unfold over time. Typically, we are interested in testing whether such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset and offset of effects (e.g., differences across conditions). Here, we introduce a novel *model-based approch* for analysing one-dimensional M/EEG timeseries such as ERPs or decoding timecourses and their differences across conditions. This approach relies on Bayesian nonparametric multilevel modelling (multilevel generalised additive models or Gaussian processes), which outputs posterior probabililities of the effect being above chance at every timestep/voxel, while naturally taking into account the temporal dependencies and between-subject variability present in such data."
keywords: [EEG, MEG, generalised additive models, gaussian processes, mixed-effects models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(grateful)
library(permuco)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)

@@

# setting a default ggplot2 theme
theme_set(theme_light(base_size = 10, base_family = "Open Sans") )

# template for citing R packages
pkrt_set(pkg = "the `R` package `:pkg` v :ver [:ref]")
```

@@

$$
\begin{aligned} 
y_{i} &\sim \mathrm{EF}\left(\mu_{i}, \phi\right)\\
g\left(\mu_i\right) &= A_{i} + \mathbf{X}_{i} \gamma + \sum_{j=1}^{J} f_{j}\left(x_{ij}\right)
\end{aligned}
$$

where $y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)$ denotes that the observations $y_{i}$ are distributed as some member of the exponential family of distributions (e.g., Gaussian, Gamma, Beta, Poisson) with mean $\mu_{i}$ and scale parameter $\phi$; $g(\cdot)$ is the link function, $A$ is an offset, $\mathbf{X}_{i}$ is the $i$th row of a parametric model matrix, $\gamma$ is a vector of parameters for the parametric terms, $f_{j}$ is a smooth function of covariate $x_{j}$. The smooth functions $f_{j}$ are represented in the model via penalised splines basis expansions of the covariates, that are a weighted sum of basis functions:

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. Splines' coefficients are penalised...

@@

A Gaussian process (GP) is a stochastic process that defines the distribution over a collection of random variables indexed by a continuous variable, that is $\{f(t): t \in \mathcal{T}\}$ for some index set $\mathcal{T}$ [@riutort-mayol_practical_2023; @rasmussen2005]. Whereas Bayesian linear regression outputs a distribution over the parameters of some predefind parametric model, the GP approach, in contrast, is a non-parametric approach, in that it finds a distribution over the possible functions that are consistent with the observed data. However, note that nonparametric does not mean there aren't parameters, it means that there are infinitely many parameters.

From [brms documentation](https://www.rdocumentation.org/packages/brms/versions/2.22.0/topics/gp): A GP is a stochastic process, which describes the relation between one or more predictors $x=\left(x_{1}, \ldots, x_{d}\right)$ and a response $f(x)$, where $d$ is the number of predictors. A GP is the generalization of the multivariate normal distribution to an infinite number of dimensions. Thus, it can be interpreted as a prior over functions. The values of $f()$ at any finite set of locations are jointly multivariate normal, with a covariance matrix defined by the covariance kernel $k_p\left(x_i, x_j\right)$, where $p$ is the vector of parameters of the GP:

$$
\left(f\left(x_{1}\right), \ldots f\left(x_{n}\right) \sim \operatorname{MVN}\left(0, \left(k_p\left(x_{i}, x_{j}\right)\right)_{i, j=1}^{n}\right)\right.
$$

The smoothness and general behaviour of the function $f$ depends only on the choice of covariance kernel, which ensures that values that are close together in the input space will be mapped to similar output values...

From this perspective, $f$ is a realisation of an infinite dimensional normal distribution:

@@

where $C$ is a covariance kernel with hyperparameters $\lambda$ that defines the covariance between two function values $f\left(t_1\right)$ and $f\left(t_2\right)$ for two time points $t_1$ and $t_2$ [@rasmussen2005]. Similar to the different choices of the basis function for splines, different choices of the covariance kernel lead to different GPs. In this article, we consider the squared-exponential (a.k.a. radial basis function) kernel, which computes the squared distance between points and converts it into a measure of similarity. It is defined as:

$$
C(\lambda) := C\left(t_1, t_2, \sigma, \gamma\right) := \sigma^2 \exp \left(-\frac{||t_1-t_2||^{2}}{2 \gamma^2}\right)
$$

with hyperparameters $\lambda = (\sigma, \gamma)$, expressing the overall scale of GP and the length-scale, respectively [@rasmussen2005]. The advantages of this kernel are that it is computationally efficient and (infinitely) smooth making it a reasonable choice for the purposes of the present article. Here again, $\lambda$ hyperparameters are estimated from the data, along with all other model parameters.

Taken from <https://michael-franke.github.io/Bayesian-Regression/practice-sheets/10c-Gaussian-processes.html>: For a given vector $\mathbf{x}$, we can use the kernel to construct finite multi-variate normal distribution associated with it like so:

$$
\mathbf{x} \mapsto_{G P} \operatorname{MVNormal}(m(\mathbf{x}), k(\mathbf{x}, \mathbf{x}))
$$

where $m$ is a function that specifies the mean for the distribution associated with $\mathbf{x}$. This mapping is essentially the Gaussian process: a systematic association of vectors of arbitrary length with a suitable multi-variate normal distribution.

@@

<!--

@@

-->

Following the approach used by @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see details and code in @yeung2004]. As in @rousselet_using_2025, the signal was generated from truncated Gaussian with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to & 500 Hz sampling rate. We simulated such data for a group of 20 participants with 50 trials per participant and condition.

@@

# importing helper functions
source("code/functions.R")

@@

```{r, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Slope for the difference between conditions according to the GAM (left) or the GP (right)."}
# check posterior_smooths()?
# https://www.rdocumentation.org/packages/brms/versions/2.22.0/topics/posterior_smooths.brmsfit
# defining a function to plot posterior slope
plot_post_slope <- function (model, data, eps = 0.1) {

@@

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = x, y = prob_ratio) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_line() +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 15,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
    labs(
        x = "Time (ms)",
        y = expression(Pr(beta>0~"|"~data, model) / (1 - Pr(beta>0~"|"~data) ) )
        )

@@

# plotting it
gp_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = x, y = prob_ratio) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_line() +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 15,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
    labs(
        x = "Time (ms)",
        y = expression(Pr(beta>0~"|"~data, model) / (1 - Pr(beta>0~"|"~data) ) )
        )

@@

```{r mne-decoding, out.width = "75%", eval = FALSE}
# loading the reticulate package
library(reticulate)

# importing the numpy and mne python modules
np <- import("numpy")
mne <- import("mne")
sklearn <- import("sklearn")

# defining the function in R (it will be executed in Python)
mne_decoding <- function (X, labels, ncores = 1) {

    # converting R dataframe to NumPy array (reshaping if needed)
    # X should be a matrix before conversion
    X_np <- np$array(X)

    if (length(dim(X_np) ) == 2) {

        # adding a second dimension (channels) if missing
        X_np <- np$expand_dims(X_np, axis = as.integer(1) )

    # defining the classifier
    clf <- sklearn$linear_model$LogisticRegression(solver = "liblinear")

    # sliding the estimator on all time frames
    time_decod <- mne$decoding$SlidingEstimator(
        clf, n_jobs = as.integer(ncores),
        scoring = "roc_auc", verbose = TRUE
        )

    # or using N-fold cross-validation
    scores <- mne$decoding$cross_val_multiscore(
        time_decod,
        X_np,
        labels,
        cv = as.integer(4),
        n_jobs = as.integer(ncores),
        verbose = TRUE
        )

    # returning the scores (averaged over CV folds)
    return (scores)

# listing all participants
participants <- unique(raw_df$participant)

# initialising empty decoding results
group_decoding_scores <- data.frame()

# running decoding for each participant
for (ppt in participants) {

    # printing progress
    print(ppt)

    # retrieve data from one participant
    ppt_data <- raw_df %>%
        filter(participant == ppt) %>%
        select(-participant) %>%
        pivot_wider(names_from = x, values_from = y) %>%
        select(-condition, -trial)

    # extracting the labels
    labels <- raw_df %>%
        filter(participant == ppt) %>%
        select(-participant) %>%
        pivot_wider(names_from = x, values_from = y) %>%
        pull(condition) %>%
        as.numeric()

    # extracting the timesteps
    timesteps <- raw_df %>%
        filter(participant == ppt) %>%
        select(-participant) %>%
        pull(x) %>%
        unique()

    # running the decoding
    decoding_scores <- data.frame(
        mne_decoding(X = ppt_data, labels = labels-1)
        ) %>%
        # computing the average over CV folds
        summarise(across(where(is.numeric), mean) )

    # appending to previous results
    group_decoding_scores <- bind_rows(group_decoding_scores, decoding_scores)

# saving the scores
saveRDS(object = group_decoding_scores, file = "results/decoding_scores.rds")
```

```{r, out.width = "75%", fig.cap = "Exemplary group-level average timecourse of binary decoding accuracy (ROC AUC). Decoding was performed using MNE-Python."}
# importing the decoding scores
group_decoding_scores <- readRDS(file = "results/decoding_scores.rds")

# extracting the timesteps
timesteps <- raw_df %>%
    # filter(participant == ppt) %>%
    # select(-participant) %>%
    pull(x) %>%
    unique()

# plotting it
group_decoding_scores %>%
    t() %>%
    data.frame() %>%
    mutate(
        mean = rowMeans(across(1:20) ),
        se = apply(across(1:20), 1, function (x) sd(x) / sqrt(length(x) ) )
        ) %>%
    mutate(lower = mean - se, upper = mean + se) %>%
    mutate(time = timesteps) %>%
    ggplot(aes(x = time, y = mean) ) +
    geom_hline(yintercept = 0.5, linetype = "dashed") +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.2) +
    geom_line(color = "steelblue", size = 0.8) +
    labs(x = "Time (s)", y = "Decoding accuracy (ROC AUC)")
```

@@

    # plotting it
    prob_y_above_0 %>%
        mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
        ggplot(aes(x = time, y = prob_ratio) ) +
        geom_hline(yintercept = 1, linetype = "dashed") +
        geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
        geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
        geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
        geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
        geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
        geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
        geom_vline(xintercept = 0, linetype = 2) +
        geom_line() +
        geom_point(
            data = . %>% dplyr::filter(!is.na(above_thres) ),
            aes(y = threshold),
            colour = "darkgreen",
            shape = 15,
            size = 3,
            na.rm = TRUE
            ) +
        scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
        labs(
            x = "Time (s)",
            y = expression(
                Pr(AUC>chance~"|"~data, model) / (1 - Pr(AUC>chance~"|"~data, model) )
                )
            )

@@

```{r fig-sim-timegen, echo = FALSE, dpi = 300, fig.cap = "Exemplary (simulated) group-level average cross-temporal generalisation matrix of decoding performance (ROC AUC)."}
# number of participants
n_participants <- 10

# number of timepoints
# n_timepoints <- 240 # (1.2s at 200Hz)
n_timepoints <- 60 # (1.2s at 50Hz)
timepoints <- seq(from = -0.2, to = 1, length.out = n_timepoints)

@@

# plotting cross-temporal generalization matrix
timegen_data %>%
    summarise(auc = mean(auc), .by = c(train_time, test_time) ) %>%
    ggplot(aes(x = train_time, y = test_time, fill = auc) ) +
    geom_tile() +
    geom_vline(xintercept = 0, linetype = 2, linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = 2, linewidth = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, linewidth = 0.5) +
    scale_x_continuous(expand = c(0, 0) ) +
    scale_y_continuous(expand = c(0, 0) ) + 
    coord_fixed() +
    scale_fill_scico(palette = "vik", midpoint = 0.5, name = "AUC") +
    labs(x = "Testing time (s)", y = "Training time (s)")
```

Now, we want to test whether and when decoding performance is above chance level (0.5 for a binary decoding task). These two models are computationally heavier to fit (more observations and 2D smooth functions)...

```{r gam-gp-timegen, eval = FALSE, message = FALSE}
# fitting a GAM with two temporal dimensions
timegen_gam <- brm(
    # 2D thin-plate spline (tp) to model smooth interactions between training and testing time
    # auc ~ s(train_time, test_time, bs = "tp", k = 10),
    auc ~ t2(train_time, test_time, bs = "tp", k = 10),
    data = timegen_data,
    family = Beta(),
    iter = 2000,
    chains = 4,
    cores = 4,
    file = "models/timegen_gam_t2.rds"
    )

# fitting a GP with two temporal dimensions
timegen_gp <- brm(
    auc ~ gp(train_time, test_time, k = 20),
    # data = timegen_data %>% mutate(auc_c = scale(x = auc, center = TRUE, scale = FALSE) ),
    data = timegen_data,
    family = Beta(),
    control = list(adapt_delta = 0.95, max_treedepth = 20),
    iter = 2000,
    chains = 4,
    cores = 4,
    file = "models/timegen_gp.rds"
    )
```

@@

@@

# checking timecourse of timegen_data's diagonal
timegen_data %>%
    filter(train_time == test_time) %>%
    summarise(auc = mean(auc), .by = train_time) %>%
    ggplot(aes(x = train_time, y = auc) ) +
    geom_hline(yintercept = 0.5, linetype = 2) +
    geom_line() +
    labs(x = "Time (s)", y = "AUC")
```

@@

## Comparing the identified onsets/offsets to other approaches

From Rousselet (2024): Five methods for multiple comparison correction were considered: two FDR methods; two cluster-based methods; and the maximum statistics. The two FDR methods were BH95 (Benjamini & Hochberg, 1995) and BY01 (Benjamini & Yekutieli, 2001), which were applied to the permutation p-values using the `p.adjust` function in `R`. The first cluster-based inference was implemented using a cluster-sum statistic of squared t-values...

Comparing performance of the GAM and GP models with:

- uncorrected univariate tests
- univariate tests + Bonferroni
- univariate tests + BH95 (Benjamini-Hochberg, 1995) aka FDR
- univariate tests + BY01 (Benjamini-Yekutieli, 2001)
- cluster-based permutation test (cluster sum)
- cluster-based permutation test (cluster depth)
- threshold-free cluster enhancement
- change point

We used `r pkrt("permuco")` and `r pkrt("changepoint")`...

```{r permutations-1d, echo = FALSE, eval = FALSE, fig.cap = "T-values timecourse with onset identified using the permuco package (clustermass method)."}
# centering AUC relative to chance level (0.5)
decoding_data_perm <- decoding_data %>%
    mutate(auc_centered = auc - 0.5) %>%
    dplyr::select(-auc)

@@

Prepare a wrapper `R` package and show how to call it in `Python` and integrate it with `MNE-Python` [@gramfort2013] pipelines...

@@

# Packages

```{r, echo = FALSE}
# citing the R packages we have used
cite_packages(output = "paragraph", omit = NULL, out.dir = ".")
```

@@

$$
\text{TFCE} = \int_{h} e(h)^{E} h^{H} \mathrm{d}h
$$
